---
title: "Getting started with Triage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Triage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Getting started with Triage

Triage is a downstream workflow for adjudicating cell type annotations
after candidate labels have been generated by separate annotation
methods. It evaluates reviewer assignments together with differential
expression, enrichment evidence, and Cell Ontology relationships to
determine which candidate identity and specificity level is supported.

The repository has two components:

1. **The R package** (`Triage`) — installable, loads without API keys,
   exposes a compact public API.
2. **`reproducibility/`** — resources for reproducing the reported
   benchmark analyses.

## 1. Installation

```{r install}
# From a local clone:
R CMD build .
R CMD INSTALL Triage_1.0.0.tar.gz

# Or from GitHub:
# remotes::install_github("shaolab-UM/Triage")
library(Triage)
```

The package loads without any API keys; keys are only needed for the
optional API path (section 7).

## 2. Loading the Cell Ontology

Triage ships the Cell Ontology JSON snapshot used by the manuscript
(v2025-07-30) in `inst/extdata/ontology/`. All ontology lookups are
local; OLS web queries are never required.

```{r ontology}
ontology <- load_triage_ontology()
```

## 3. Reading reviewer inputs

Reviewer input dossiers for the released benchmark are assembled by the
stage-08 pipeline script (`reproducibility/scripts/pipeline/08_build_judge_inputs.R`);
the package-side reader parses and validates them:

```{r input}
jin <- read_triage_input("examples/census_immune_cluster1/cluster_1_round1.json")
```

`read_triage_input()` accepts a file path, a JSON string, or a parsed
list, and validates the required `cluster_id` / `inputs` fields.

## 4. CL-Linker: deterministic label mapping

CL-Linker maps free-text reviewer labels to Cell Ontology terms:

```{r cl_linker}
run_cl_linker(c("macrophage", "T cell"), c("", "CL:0000084"))
```

## 5. Assembling an adjudication input

```{r build}
ain <- build_adjudication_input(
  cluster_id = "cluster_1",
  reviewers = list(
    cassia = list(top1_cell_type = "classical monocyte",
                  topk_cell_types = c("classical monocyte",
                                      "intermediate monocyte")),
    in_house = list(top1_cell_type = "monocyte",
                    cell_ontology_id = "CL:0000576"))
)
```

## 6. Running an adjudication

Two modes:

- **Deterministic postprocessing:** set `use_api = FALSE` and supply a
  precomputed Handling Editor output; the package executes only the
  deterministic downstream chain (release policy, CL normalization,
  local gate). The Handling Editor is NOT executed.
- **Handling Editor API:** set `use_api = TRUE` and supply an API key and
  a chat-completions endpoint through `api_key` / `api_base_url` (or via
  the `DEEPSEEK_API_KEY` / `LLM_API_BASE_URL` environment variables); the
  optional Handling Editor API path runs, then the same deterministic
  chain. The manuscript workflow scripts are under
  `reproducibility/scripts/`.

```{r adjudication}
# Deterministic example (see examples/census_immune_cluster1/):
fin <- run_triage_adjudication(
  jin,
  ontology = ontology,
  use_api = FALSE,
  head_output = "examples/census_immune_cluster1/head_round1.json",
  dataset_name = "census_immune",
  project_root = "reproducibility/primary"
)
attr(fin, "gate")$ok

# API path (uses the package default model):
# fin <- run_triage_adjudication(jin, ontology = ontology,
#          use_api = TRUE, temperature = 0, prompt_profile = "compact",
#          api_key = "YOUR_API_KEY",
#          api_base_url = "https://api.deepseek.com/chat/completions")
# api_key / api_base_url can be omitted when the DEEPSEEK_API_KEY and
# LLM_API_BASE_URL environment variables are set.
```

## 7. Validating results

```{r validate}
validate_triage_result(fin, jin)
```

`validate_triage_result()` re-runs the deterministic local gate
(schema, banned tokens, citation policy, reviewer-support logic) and
stops with the failing reasons if anything is wrong.

## 8. Reproducing the manuscript

See `reproducibility/README.md` and `docs/SELECTOR_REPRODUCTION.md`,
`docs/CL_LINKER_EVALUATION.md`, `docs/CELLMARKERDB_SENSITIVITY.md`.
All release validation scripts are deterministic (no LLM calls) and live
in `reproducibility/scripts/`; the manuscript workflow scripts are
separate from the package API.
