---
title: "Getting started with Triage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Triage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Getting started with Triage

Triage is a downstream workflow for adjudicating cell type annotations
after candidate labels have been generated by separate annotation
methods. It evaluates reviewer assignments together with differential
expression, enrichment evidence, and Cell Ontology relationships to
determine which candidate identity and specificity level is supported.

The repository has two components:

1. **The R package** (`Triage`) — installable, loads without API keys,
   exposes a compact public API. This vignette covers the package only;
   the `reproducibility/` tree is not installed with the package.
2. **`reproducibility/`** — resources for reproducing the reported
   benchmark analyses (available in the GitHub repository checkout).

## 1. Installation

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("shaolab-UM/Triage")
library(Triage)
```

The package loads without any API keys; keys are only needed for the
optional API path (section 6).

Triage provides three distinct entry points, in increasing scope:

- **A. Deterministic quick start** — `run_triage_example()`: runs the
  packaged, fully deterministic example end-to-end. No API key and no
  repository checkout required (section 3).
- **B. Lightweight adjudication API** — `run_triage_adjudication()` for a
  single cluster (sections 4-5).
- **C. Full DEG-input workflow** — `run_triage()`: the complete
  multi-stage workflow starting from a cluster-level DEG/marker table
  (section 7). Repository checkout is only needed for manuscript
  reproduction (section 9), not for any of these entry points.

## 2. Loading the Cell Ontology

Triage ships the Cell Ontology JSON snapshot used by the manuscript
(v2025-07-30) in `inst/extdata/ontology/`. All ontology lookups are
local; OLS web queries are never required.

```{r ontology}
ontology <- load_triage_ontology()
```

## 3. Deterministic quick start

```{r quickstart}
fin_example <- run_triage_example()
```

`run_triage_example()` loads the bundled Cell Ontology, reads reviewer
fixtures shipped inside the installed package
(`system.file("extdata", "examples", "census_immune_cluster1", ...)`),
executes the deterministic adjudication chain, validates the result, and
returns it invisibly. No API key, network call, or repository checkout is
involved.

## 4. Reading reviewer inputs

Reviewer input dossiers for the released benchmark are assembled by the
stage-08 pipeline script of the manuscript workflow (repository
checkout); the package-side reader parses and validates them:

```{r input}
input_path <- system.file("extdata", "examples", "census_immune_cluster1",
                          "cluster_1_round1.json", package = "Triage")
jin <- read_triage_input(input_path)
```

`read_triage_input()` accepts a file path, a JSON string, or a parsed
list, and validates the required `cluster_id` / `inputs` fields.

## 5. CL-Linker: deterministic label mapping

CL-Linker maps free-text reviewer labels to Cell Ontology terms:

```{r cl_linker}
run_cl_linker(c("macrophage", "T cell"), c("", "CL:0000084"))
```

## 6. Building and running an adjudication

```{r build}
ain <- build_adjudication_input(
  cluster_id = "cluster_1",
  reviewers = list(
    cassia = list(top1_cell_type = "classical monocyte",
                  topk_cell_types = c("classical monocyte",
                                      "intermediate monocyte")),
    in_house = list(top1_cell_type = "monocyte",
                    cell_ontology_id = "CL:0000576"))
)
```

## 6. Running an adjudication

Three convenient entry points:

- **Packaged example:** `run_triage_example()` runs the full deterministic
  example below end-to-end, using fixtures shipped inside the installed
  package. No API key and no repository checkout required.
- **Deterministic postprocessing:** set `use_api = FALSE` and supply a
  precomputed Handling Editor output; the package executes only the
  deterministic downstream chain (release policy, CL normalization,
  local gate). The Handling Editor is NOT executed.
- **Handling Editor API:** set `use_api = TRUE` and supply an API key and
  a chat-completions endpoint through `api_key` / `api_base_url` (or via
  the `DEEPSEEK_API_KEY` / `LLM_API_BASE_URL` environment variables); the
  optional Handling Editor API path runs, then the same deterministic
  chain. The manuscript workflow scripts are separate from the package
  API (repository checkout).

```{r adjudication, eval=FALSE}
# Deterministic example with packaged fixtures:
fin <- run_triage_adjudication(
  jin,
  ontology = ontology,
  use_api = FALSE,
  head_output = system.file("extdata", "examples", "census_immune_cluster1",
                            "head_round1.json", package = "Triage"),
  dataset_name = "census_immune",
  project_root = tempdir()
)
attr(fin, "gate")$ok

# API path (uses the package default model):
# fin <- run_triage_adjudication(jin, ontology = ontology,
#          use_api = TRUE, temperature = 0, prompt_profile = "compact",
#          api_key = "YOUR_API_KEY",
#          api_base_url = "https://api.deepseek.com/chat/completions")
# api_key / api_base_url can be omitted when the DEEPSEEK_API_KEY and
# LLM_API_BASE_URL environment variables are set.
```

## 7. Full DEG-input workflow (R-only)

`run_triage()` runs the complete multi-stage workflow from a
cluster-level DEG/marker table (columns: `cluster`, `gene`, `avg_log2FC`,
`p_val`, `p_val_adj`, `pct.1`, `pct.2`), through upstream LLM-backed
reviewer stages to the final adjudication, entirely from R (no repository
checkout, no shell scripts):

```{r fullworkflow, eval=FALSE}
# One-time per machine:
install_triage_dependencies(species = "human")
resources <- setup_triage_resources(species = "human")   # downloads STRING
                                                          # + CollecTRI
# Per run:
triage_preflight(species = "human", deg = "path/to/markers.csv",
                 api_key = "YOUR_API_KEY",
                 api_base_url = "https://api.deepseek.com/chat/completions")
res <- run_triage(deg = "path/to/markers.csv", species = "human",
                  tissue = "pancreas", out = "results")
res$summary
```

Original cluster identifiers are deterministically anonymized to
`cluster_1...N` before any stage runs (the mapping is kept locally in
`cluster_map.csv`); reference labels are never used for adjudication and
are only consumed when explicitly supplied for optional evaluation.
API keys and a chat-completions endpoint can be passed as arguments or
through the `DEEPSEEK_API_KEY` / `LLM_API_BASE_URL` environment
variables.

## 8. Validating results

```{r validate}
validate_triage_result(fin, jin)
```

`validate_triage_result()` re-runs the deterministic local gate
(schema, banned tokens, citation policy, reviewer-support logic) and
stops with the failing reasons if anything is wrong.

## 9. Reproducing the manuscript

Repository checkout is required only for manuscript reproduction. The
reproduction resources (data, deterministic release checks, and
manuscript workflow scripts) live under `reproducibility/` in the GitHub
repository checkout; see its README plus `docs/SELECTOR_REPRODUCTION.md`,
`docs/CL_LINKER_EVALUATION.md`, and `docs/CELLMARKERDB_SENSITIVITY.md`.
All release validation scripts are deterministic (no LLM calls) and the
manuscript workflow scripts are separate from the package API.
