---
title: "Getting started with Triage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Triage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Getting started with Triage

Triage is an evidence-adjudication pipeline for single-cell cluster
annotation. It combines three independent reviewers — a CASSIA agent,
an in-house marker-based annotator, and an enrichment reviewer
(clusterProfiler) — with a deterministic Cell-Ontology (CL) mapping
layer and an LLM handling editor / chief-QC step.

The repository has two components:

1. **The R package** (`Triage`) — installable, loads without API keys,
   exposes a compact public API.
2. **`reproducibility/`** — manuscript data, scripts, and expected
   results for exact reproduction.

## 1. Installation

```{r install}
# From a local clone:
R CMD build .
R CMD INSTALL Triage_1.0.0.tar.gz

# Or from GitHub (once pushed):
# remotes::install_github("shaolab-UM/Triage")
library(Triage)
```

The package loads without any API keys; keys are only needed for the
optional API path (section 7).

## 2. Loading the Cell Ontology

Triage ships the Cell Ontology JSON snapshot used by the manuscript
(v2025-07-30) in `inst/extdata/ontology/`. All ontology lookups are
local; OLS web queries are never required.

```{r ontology}
ontology <- load_triage_ontology()
default_ontology_path()   # honors the CL_LOCAL_JSON env override
```

## 3. Reading reviewer inputs

```{r input}
jin <- read_triage_input("reproducibility/primary/Census_immune/handling_editor_round1_inputs/cluster_1_round1.json")
```

`read_triage_input()` accepts a file path, a JSON string, or a parsed
list, and validates the required `cluster_id` / `inputs` fields.

## 4. Mapping labels to the Cell Ontology

```{r mapping}
map_cell_ontology("monocyte", ontology = ontology)
```

## 5. CL-Linker: deterministic label mapping

```{r cl_linker}
run_cl_linker(c("macrophage", "T cell"), c("", "CL:0000084"))
```

## 6. Reviewer summaries and adjudication inputs

```{r build}
rs <- build_reviewer_summary(list(
  cassia = list(top1_cell_type = "classical monocyte",
                topk_cell_types = c("classical monocyte",
                                    "intermediate monocyte")),
  in_house = list(top1_cell_type = "monocyte",
                  cell_ontology_id = "CL:0000576")
))

ain <- build_adjudication_input(
  cluster_id = "cluster_1",
  reviewers = rs
)
```

## 7. Running an adjudication

Two modes:

- **No API (deterministic):** supply a precomputed head-editor output;
  Triage runs the full deterministic post-processing chain
  (release policy, CL normalization, local gate).
- **API:** set `use_api = TRUE` with `DEEPSEEK_API_KEY` and
  (optionally) `LLM_API_BASE_URL`; the head editor model is called
  first, then the same deterministic chain runs.

```{r adjudication}
# Deterministic example (see examples/census_immune_cluster1_demo/):
fin <- run_triage_adjudication(
  jin,
  ontology = ontology,
  use_api = FALSE,
  head_output = "examples/census_immune_cluster1_demo/head_round1.json",
  dataset_name = "census_immune",
  project_root = "reproducibility/primary"
)
attr(fin, "gate")$ok

# API path (primary workflow: all roles deepseek-v4-flash, temp 0):
# fin <- run_triage_adjudication(jin, ontology = ontology,
#          use_api = TRUE, model = "deepseek-v4-flash",
#          temperature = 0, prompt_profile = "compact")
```

## 8. Validating results

```{r validate}
validate_triage_result(fin, jin)
```

`validate_triage_result()` re-runs the deterministic local gate
(schema, banned tokens, citation policy, reviewer-support logic) and
stops with the failing reasons if anything is wrong.

## 9. Reproducing the manuscript

See `README.md` ("Reproducing the manuscript analyses") and
`docs/SELECTOR_REPRODUCTION.md`, `docs/CL_LINKER_EVALUATION.md`,
`docs/CELLMARKERDB_SENSITIVITY.md`. All release validation scripts are
deterministic (no LLM calls) and live in `reproducibility/scripts/`.
