---
title: "Getting started with Triage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Triage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Getting started with Triage

Triage is a downstream workflow for adjudicating cell type annotations
after candidate labels have been generated by separate annotation
methods. It evaluates reviewer assignments together with differential
expression, enrichment evidence, and Cell Ontology relationships to
determine which candidate identity and specificity level is supported.

The repository has two components:

1. **The R package** (`Triage`) — installable, loads without API keys,
   exposes a compact public API. This vignette covers the package only;
   the `reproducibility/` tree is not installed with the package.
2. **`reproducibility/`** — resources for reproducing the reported
   benchmark analyses (available in the GitHub repository checkout).

## 1. Installation

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("shaolab-UM/Triage")
library(Triage)
```

The package loads without any API keys; keys are only needed for the
optional API path (section 6).

## 2. Loading the Cell Ontology

Triage ships the Cell Ontology JSON snapshot used by the manuscript
(v2025-07-30) in `inst/extdata/ontology/`. All ontology lookups are
local; OLS web queries are never required.

```{r ontology}
ontology <- load_triage_ontology()
```

## 3. Reading reviewer inputs

Reviewer input dossiers for the released benchmark are assembled by the
stage-08 pipeline script of the manuscript workflow (repository
checkout); the package-side reader parses and validates them:

```{r input}
input_path <- system.file("extdata", "examples", "census_immune_cluster1",
                          "cluster_1_round1.json", package = "Triage")
jin <- read_triage_input(input_path)
```

`read_triage_input()` accepts a file path, a JSON string, or a parsed
list, and validates the required `cluster_id` / `inputs` fields.

## 4. CL-Linker: deterministic label mapping

CL-Linker maps free-text reviewer labels to Cell Ontology terms:

```{r cl_linker}
run_cl_linker(c("macrophage", "T cell"), c("", "CL:0000084"))
```

## 5. Assembling an adjudication input

```{r build}
ain <- build_adjudication_input(
  cluster_id = "cluster_1",
  reviewers = list(
    cassia = list(top1_cell_type = "classical monocyte",
                  topk_cell_types = c("classical monocyte",
                                      "intermediate monocyte")),
    in_house = list(top1_cell_type = "monocyte",
                    cell_ontology_id = "CL:0000576"))
)
```

## 6. Running an adjudication

Three convenient entry points:

- **Packaged example:** `run_triage_example()` runs the full deterministic
  example below end-to-end, using fixtures shipped inside the installed
  package. No API key and no repository checkout required.
- **Deterministic postprocessing:** set `use_api = FALSE` and supply a
  precomputed Handling Editor output; the package executes only the
  deterministic downstream chain (release policy, CL normalization,
  local gate). The Handling Editor is NOT executed.
- **Handling Editor API:** set `use_api = TRUE` and supply an API key and
  a chat-completions endpoint through `api_key` / `api_base_url` (or via
  the `DEEPSEEK_API_KEY` / `LLM_API_BASE_URL` environment variables); the
  optional Handling Editor API path runs, then the same deterministic
  chain. The manuscript workflow scripts are separate from the package
  API (repository checkout).

```{r adjudication, eval=FALSE}
# Deterministic example with packaged fixtures:
fin <- run_triage_adjudication(
  jin,
  ontology = ontology,
  use_api = FALSE,
  head_output = system.file("extdata", "examples", "census_immune_cluster1",
                            "head_round1.json", package = "Triage"),
  dataset_name = "census_immune",
  project_root = tempdir()
)
attr(fin, "gate")$ok

# API path (uses the package default model):
# fin <- run_triage_adjudication(jin, ontology = ontology,
#          use_api = TRUE, temperature = 0, prompt_profile = "compact",
#          api_key = "YOUR_API_KEY",
#          api_base_url = "https://api.deepseek.com/chat/completions")
# api_key / api_base_url can be omitted when the DEEPSEEK_API_KEY and
# LLM_API_BASE_URL environment variables are set.
```

## 7. Validating results

```{r validate}
validate_triage_result(fin, jin)
```

`validate_triage_result()` re-runs the deterministic local gate
(schema, banned tokens, citation policy, reviewer-support logic) and
stops with the failing reasons if anything is wrong.

## 8. Reproducing the manuscript

The reproduction resources (data, deterministic release checks, and
manuscript workflow scripts) live in the GitHub repository checkout under
`reproducibility/`; see its README plus `docs/SELECTOR_REPRODUCTION.md`,
`docs/CL_LINKER_EVALUATION.md`, and `docs/CELLMARKERDB_SENSITIVITY.md`.
All release validation scripts are deterministic (no LLM calls) and the
manuscript workflow scripts are separate from the package API.
